# Chapter 1 - Brief introduction

Hello, I am Juho Mononen, a PhD student from UEF who who has *some* experience with Github and is quite familiar with R and Rstudio. Currently dappling in bioinformatics with a main focus of identifying regulatory sites affected by genetic variation. I am mostly self taught in programming, so participating to unlearn potentially bad habits and to learn better practices with project managing and statistics. 

A friend introduced me to this course.

## Things  I expect to learn:
* Basics of version control and Github usage with R
* Good practices with R and Rstudio
 
## Links

### Repo
**_[Link to my Github](https://github.com/juhomon/IODS-project)_**

```{r, fig.height=3, fig.width=3}
# Smiley 

## Somewhat disturbing smiley for testing
plot(x = c(-0.5,0.5), y = c(6,6), ylim=c(0,8), xlim = c(-1,1), cex = 5, xlab = "" , ylab = "", axes = F, pch = 20, col = "dodgerblue4")
curve(-log(1-x^2), from = -1, to = 1, add = T, col = "dodgerblue4", lwd = 5)

```

```{r}
date()
```